The error in Bayesian phylogenetic reconstruction when speciation co-occurs

### 2018-09-27 article.tex
Write v0.5, fix #11

\gio{I've taken the freedom to change your proposed structure. Feel free to change whatever you don't like it. I might be not completely accurate on the biological background.}
\richel{It's just fine :-)}

Current phylogenetic tools assume that only a single speciation event can occur at the same time.
While this assumption is useful to construct a wide variety of successful models \richel{@gio: citation here},
they disallow for environmental changes that trigger speciation on a large scale,
for example, the cichlid fish diversification in the 
African Great Lakes: Malawi, Tanganyika and Victoria \richel{@gio: citation here}.

@@

\gio{If I described the process in the same way you report in the example I would probably end up writing the same things that we say a few lines below, where we describe the parameters. Don't you think?}
\richel{You described the model in the Methods. I moved it to here}
In the MBD model, parameters $\lambda$ and $\mu$ correspond, respectively, 
to the usual per-species speciation and extinction rates. 
The model also introduces two new parameters: 
$\nu$ is the total rate for an environmental change to trigger, 
which leads to a potentially multiple speciation event 
\richel{How exactly? When the environment change is triggered, is the speciation rate $q$ or $\lambda + q$? What ends the change?}.
If such event triggers each species present at that moment in time can undergo a speciation event with probability $q$.

@@

The MBD model has 4 parameters, depicted in table \ref{table:parameters}.

\richel{describe parameter values used here, example is below}
\gio{@richel: I described the meaning of each parameter. I guess it is fine in this way. For the setting I think we have to wait to decide which values are actually worth a try}
\richel{I assume you have some idea of parameter values? For example, I think you will not use a $\lambda$ of -3 or 1234. Please put some reasonable values here, like I did in Figure 1 for you}
% The per species speciation completion rates $\lambda$ we use are $0.1$, $0.3$, $1.0$ and $10^9$. This means that there is a $0.15 dt$ ($0.35 dt$, $1.0 dt$, $10^9 dt$) probability of speciation completion occurring in an infinitesimal time $dt$. 
% We use per species extinction rates of $\mu = \mu_g = \mu_i$ $0.0$, $0.1$ and $0.2$.
% For each combination of $\lambda$, $\mu$ and $n$, we use a speciation initiation rate $b = b_i = b_g$ that gives a given value of the expected number of species. We use expected mean tree sizes $n$ of 50, 100 and 200 good taxa and the corresponding speciation initiation rates for all parameter combinations are shown in table \ref{table:pbd_parameters}.
% Our parameters are inspired by existing work on protracted speciation \cite{etienne_and_rosindell_2012}\cite{etienne2014}.
We use \richel{please check me} $\nu = 0$ as our control for which the MBD model reduces to the BD model.

We simulate protracted birth-death trees, using the MBD package (\cite{pbd}) in the R programming language (\cite{r}).
The first tree has a random number generator seed of 1, which is incremented by 1 for each simulated tree.
For each combination of $\lambda, \mu, \nu$ and $q$, 
we generate species trees with a crown age of 15 million years 
\gio{are we sure we wanna try 15 million years. so far i've been trying only with 10 million years, which most of the time is working really well}
\richel{I suggest yes. Most papers associated with Rampal use 15 million years, including my raket paper. Just use lower speciation rates}.
Only trees with the desired number of good taxa are kept.

We create one data set to explore parameter space, 
All the trees with the correct number of good species are kept.
Based on the species tree, we simulate a DNA alignment that has the same history
as this species tree, using the \verb;phangorn; package (\cite{phangorn}). 
We set the nucleotides of the DNA alignment to follow a Jukes-Cantor (\cite{jc69})
nucleotide substitution model, in which all nucleotide-to-nucleotide transitions
are equally likely. 
The DNA sequence of the root ancestor consists of four equally sized single-nucleotide 
blocks of adenine, cytosine, guanine and thymine respectively. 
For example, for a DNA sequence length of 12, this would be AAACCCGGGTTT. 
The order of nucletides does not matter in this study, 
because we do not consider several partitions of the sequence with their own parameters. 
Only the frequency of occurrence matters.
In our Bayesian inference (see below) we use the same site model as the (obviously correct) site model prior,
but we also explore the effect of assuming a more complex site model prior.
We predict with the more complex substitution model, 
that there will be more noise and hence our inference error will increase.
On the other hand, we dare not rule out that the inference error will decrease,
due to more flexibility in the more complex prior.
We set the mutation rate in such a way to maximize the information contained in the alignment.
To do so, we set the mutation rate such that we expect on average one (possibly silent) mutation per nucleotide
between crown age and present, which equates to $\frac{1}{15}$ mutations
per million years.
The DNA sequence length is chosen to provide a
resolution of $10^3$ years, 
% If sims take too long, we will decrease this resolution
that is, to have one expected nucleotide change 
per $10^3$ years per lineage on average. As one nucleotide is expected 
to have on average one (possibly silent) mutation per 15 million years, $15 \cdot 10^3$
nucleotides result in 1 mutation per alignment per $10^3$ years (which is
coincidentally the same as \cite{moller2018}). 
The simulation of these DNA aligments follows a strict clock model, 
which we will specify as one of the two clock models assumed in the Bayesian inference (see below).

\richel{must rewrite, use pirouette as a starting point}
From an alignment, we run a Bayesian analysis and 
create a posterior distribution of trees and parameters
using the \verb;pirouette; (\cite{pirouette}) package
that sets the input parameters similar to BEAUti 2 and then runs BEAST2. 
For our site model, we assume either a Jukes-Cantor or GTR nucleotide substitution model.
The Jukes-Cantor model is the correct one, as it is used for simulating that alignment,
where the GTR model is the site model that is picked as a default by most users.
For our clock model, we assume either a strict or relaxed log-normal 
clock model. 
Also here, the strict clock model is the correct one, as it is used for simulating the alignment,
but the relaxed log-normal clock model is the one most commonly used.
We set the BD model as a tree prior, 
as gauging the effect of this incorrect assumption is the goal of this study. 
We assume an MRCA prior with a tight normal distribution
around the crown age, by choosing the crown age as mean, and a standard deviation 
of $0.5 \cdot 10^{-3}$ time units,
resulting in 95\% of the crown ages inferred have the same resolution (of $10^{-3}$ time 
units) as the alignment. 
We ran the MCMC chain to generate 1111 states,
of which we remove the first 10\% (also called the 'burn-in'). 
Of the remaining
1000 MCMC states, the effective sample size (ESS) of the posterior 
must at least be 200
for a strong enough inference (\cite{beastbook}). An ESS can be increased by increasing
the number of samples or decreasing the autocorrelation between samples. 
If the ESS is less than 200, we decrease autocorrelation by doubling 
the MCMC sampling interval of that simulation, until the ESS exceeds 200.

@@

The input trees generated with a \richel{@gio: parameter that is set to reduce the MBD model to BD} 
allow us to measure the noise of the experiment.

@@

For both data sets, we display the nLTT statistics distribution per
biological parameter combination as a violin plot.
We only show the nLTT distributions
that were generated under the (correct) assumptions of a Jukes-Cantor site model
and a strict clock model, separated per sampling method used. 
We display the nLTT statistic distributions separated per site or clock model 
in the supplementary information.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{table}
  \centering 
  \begin{tabular}{p{0.05\textwidth} p{0.6\textwidth} p{0.2\textwidth}}
    \hline
                          & Description & Values \\
    \hline
    \hline
    $\lambda$             & \richel{@gio: MBD params here} & 0.1, 0.3, 1.0, $10^9$ \\
    $\lambda$             & Speciation rate & 0.1, 0.3, 1.0, $10^9$ \\
    $\mu$                 & Extinction rate & 0.0, 0.1, 0.2 \\
    \hline
    $n$                   & Number of good taxa & 50, 100, 200 \\
    $t_c$                 & Crown age & 15 \\
    $\sigma_c$            & Standard deviation around crown age & 0.001 \\
    $M_s$                 & Sampling method & S, L, R \\
    $M_c$                 & Clock model & S, RLN \\
    $M_t$                 & Site model & JC69, GTR \\
    $r$                   & Mutation rate & $\frac{1}{15}$ \\
    $l_a$                 & DNA alignment length & $15K$ \\
    $f_i$                 & MCMC sampling interval & 1K or more \\
    $R_i$                 & RNG seed incipient tree and randomly sampled species tree & 1, 2, etc. \\
    $R_a$                 & RNG seed alignment simulation & $R_i$ \\
    $R_b$                 & RNG seed BEAST2 & $R_i$ \\
    \hline
  \end{tabular}
  \caption{
    Overview of the simulation parameters. Above the horizontal line is 
    the biological parameter set. 
    The RNG seed $R_i$ is 1 for the first simulation, 2 for the next,
    and so on.  
    The clock models are abbreviated as 'S' for a strict and 'RLN' for a relaxed log-normal model.
    The site models are abbreviated as 'JC69' for Jukes-Cantor (\cite{jc69}) and 'GTR' for the generalized 
    time-reversible model (\cite{gtr}).
  }
  \label{table:parameters}
\end{table}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

\richel{put this section here, as the journal does not request for this}
RSE \richel{@gio: I assume this this is true?} conceived the idea for this experiment.
GL created and tested the MBD package.
RJCB created and tested the experiment.
GL and RJCB wrote the first draft of the manuscript. 
RSE contributed substantially to revisions.

>>>>

Current phylogenetic tools assume that only a single speciation event can occur at the same time.
While this assumption is useful to construct a wide variety of successful models \richel{@gio: citation here} \gio{@richel: basically all the models we know are based on this assumpion: DDD, PBD, BISSE, MUSSE, SECSSE, any other SSE, DAISIE etc. etc. It's a very very general feature. Maybe being specific could lead the reader to consider things that are, in the end, not essential to the story we want to tell here. Do you still think we need it?},
\richel{Yes, I think here would be a fine spot to cite some of those models, 
I think BiSSE, DAISIE, DDD and PBD would be appropriate} 
they disallow for environmental changes that trigger speciation on a large scale,
for example, the cichlid fish diversification in the 
African Great Lakes: Malawi, Tanganyika and Victoria \richel{@gio: citation here}.

@@

\gio{If I described the process in the same way you report in the example I would probably end up writing the same things that we say a few lines below, where we describe the parameters. Don't you think?}
\richel{You described the model in the Methods. I moved it to here}
In the MBD model, parameters $\lambda$ and $\mu$ correspond, respectively, 
to the usual per-species speciation and extinction rates. 
Additionally, $\nu$ is the rate at which an environmental change is triggered.
When that event is triggered, all species at that moment
have a probability $q$ to speciatiate (independent on $\lambda$).
The number of species that speciate due to this can also be zero. 
\richel{Is this correct?}.

@@

The MBD model has 4 parameters, depicted in table \ref{table:parameters}. 
We pick values of $\nu$ in such a way we expect a multiple speciation
event to be triggered zero ($\nu = 0$), once, twice, four and eight times 
\richel{I assume you can calculate the correct $\nu$}.
We pick values of $q$ that are 0.0 (a speciation barrier at the
triggered event), 0.25, 0.5 and 1.0.
We set our extinction rate $\mu$ to 0.1 in all simulation.
As we select our phylogenies on their number of lineages,
we calculate $\lambda$ in a such a way that the mean expected number
of lineages equals the desired numbers of taxa of 50, 100 and 200. 
For $\nu = 0$, the model falls back to a standard BD model.
Note that the $\lambda$ and $q$ have different units
and it is a misconception to think that for $\lambda = q$ (already
impossible due to their units) the MBD model would reduce to a BD model.

We simulate protracted birth-death trees, using the MBD package (\cite{mbd}) in the R programming language (\cite{r}).
The first tree has a random number generator seed of 1, which is incremented by 1 for each simulated tree.
For each combination of $\lambda, \mu, \nu$ and $q$, 
we generate species trees with a crown age of 15 million years 
\gio{In general [15 million years] is ok for me. Keep in mind, though, that allowing multiple speciations may lead to an explosion in the number of species. Increasing the time by a factor of n usually means increasing the expected number of species at the present by a factor proportional to $e^n$}
\richel{I know}.
Only trees with the desired number of good taxa are kept.

From an (MBD) species tree, we create a BEAST2 posterior using
the 'pirouette' (\cite{pirouette}) R package: 
'pirouette' first simulates a DNA alignment that has the same history
as the species tree, using the \verb;phangorn; package (\cite{phangorn}).
The DNA sequence of the root ancestor consists of four equally sized single-nucleotide 
blocks of adenine, cytosine, guanine and thymine 
respectively (for example, for a DNA sequence length of 12, this would 
be AAACCCGGGTTT). Throughout evolutionary time, we use equal 
mutation rates between the four DNA nucleotides, 
also called the Jukes-Cantor (\cite{jc69}) nucleotide substitution model.
The neat seperation of the nucleotides is for visualization and debugging
purposes and has no effect in any other way. 
The equal amount of nucleotides does matter, assuring any nucleotide
mutation is equally likely to be observed.

In our Bayesian inference (see below) we use the same site model as the (obviously correct) site model prior,
but we also explore the effect of assuming a more complex site model prior.
We predict with the more complex substitution model, 
that there will be more noise and hence our inference error will increase.
On the other hand, we dare not rule out that the inference error will decrease,
due to more flexibility in the more complex prior.
We set the mutation rate in such a way to maximize the information contained in the alignment.
To do so, we set the mutation rate such that we expect on average one (possibly silent) mutation per nucleotide
between crown age and present, which equates to $\frac{1}{15}$ mutations
per million years.
The DNA sequence length is chosen to provide a
resolution of $10^3$ years, 
% If sims take too long, we will decrease this resolution
that is, to have one expected nucleotide change 
per $10^3$ years per lineage on average. As one nucleotide is expected 
to have on average one (possibly silent) mutation per 15 million years, $15 \cdot 10^3$
nucleotides result in 1 mutation per alignment per $10^3$ years (which is
coincidentally the same as \cite{moller2018}). 
The simulation of these DNA aligments follows a strict clock model, 
which we will specify as one of the two clock models assumed in the Bayesian inference (see below).

From here, the 'babette' R package (\cite{babette})
takes over and converts the DNA alignment to a BEAST2 posterior.
We set up the BEAST2 analysis to assume either a Jukes-Cantor or GTR nucleotide substitution model.
The Jukes-Cantor model is the correct one, as it is used for simulating that alignment,
where the GTR model is the site model that is picked as a default by most users.
For our clock model, we assume either a strict or relaxed log-normal 
clock model. 
Also here, the strict clock model is the correct one, as it is used for simulating the alignment,
but the relaxed log-normal clock model is the one most commonly used.
We set the BD model as a tree prior, 
as gauging the effect of this incorrect assumption is the goal of this study. 
We assume an MRCA prior with a tight normal distribution
around the crown age, by choosing the crown age as mean, and a standard deviation 
of $0.5 \cdot 10^{-3}$ time units,
resulting in 95\% of the crown ages inferred have the same resolution (of $10^{-3}$ time 
units) as the alignment. 
We ran the MCMC chain to generate 1111 states,
of which we remove the first 10\% (also called the 'burn-in'). 
Of the remaining
1000 MCMC states, the Effective Sample Size (ESS) of the posterior 
must at least be 200
for a strong enough inference (\cite{beastbook}). An ESS can be increased by increasing
the number of samples or decreasing the autocorrelation between samples. 
If the ESS is less than 200, we decrease autocorrelation by doubling 
the MCMC sampling interval of that simulation, until the ESS exceeds 200.

@@

The input trees generated with a $\nu = 0$, 
in which all BEAST2's
assumptions are met,
allow us to measure the noise of the experiment.

@@

We display the data set as an nLTT statistics distribution per
parameter combination as a violin plot.
We only show the nLTT distributions
that were generated under the (correct) assumptions of a Jukes-Cantor site model
and a strict clock model, separated per sampling method used. 
We display the nLTT statistic distributions separated per site or clock model 
in the supplementary information.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{table}
  \centering 
  \begin{tabular}{p{0.05\textwidth} p{0.6\textwidth} p{0.2\textwidth}}
    \hline
                          & Description & Values \\
    \hline
    \hline
    $\lambda$             & Per-species speciation rate & calculated \\
    $\mu$                 & Per-species extinction rate & 0.0, 0.1 \\
    $\nu$                 & Multiple speciation trigger rate & occurs never, once, twice, four and eight times \\
    $q$                   & Per-species probability of multiple speciation & 0, 0.25, 0.5, 1.0 \\
    \hline
    \hline
    $n$                   & Number of good taxa & 50, 100, 200 \\
    $t_c$                 & Crown age & 15 \\
    $\sigma_c$            & Standard deviation around crown age & 0.001 \\
    $M_c$                 & Clock model & S, RLN \\
    $M_t$                 & Site model & JC69, GTR \\
    $r$                   & Mutation rate & $\frac{1}{15}$ \\
    $l_a$                 & DNA alignment length & $15K$ \\
    $f_i$                 & MCMC sampling interval & 1K or more \\
    $R_i$                 & RNG seed MBD tree generation & 1, 2, etc. \\
    $R_a$                 & RNG seed alignment simulation & $R_i$ \\
    $R_b$                 & RNG seed BEAST2 & $R_i$ \\
    \hline
  \end{tabular}
  \caption{
    Overview of the simulation parameters. Above the horizontal line are 
    the MBD model's parameters. 
    The RNG seed $R_i$ is 1 for the first simulation, 2 for the next,
    and so on.  
    The clock models are abbreviated as 'S' for a strict and 'RLN' for a relaxed log-normal model.
    The site models are abbreviated as 'JC69' for Jukes-Cantor (\cite{jc69}) and 'GTR' for the generalized 
    time-reversible model (\cite{gtr}).
  }
  \label{table:parameters}
\end{table}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

\richel{put this section here, as the journal does not request for this}
RSE conceived the idea for this experiment.
GL created and tested the MBD package.
RJCB created and tested the experiment.
GL and RJCB wrote the first draft of the manuscript. 
RSE contributed substantially to revisions.
